---
title: "Modeling demographic effects of ability"
author: "Mike"
date: "11/12/2019"
output: 
  html_document:
    toc: true
---

# Loading data 

Package loading.

```{r, message = FALSE, warn = FALSE}
library(tidyverse)
library(brms)
library(wordbankr)
theme_set(langcog::theme_mikabr())
knitr::opts_chunk$set(cache = TRUE, warn = FALSE, message = FALSE)
```

Data loading.

```{r}
wg_admins <- get_administration_data(language = "English (American)", 
                                     form = "WS", original_ids = TRUE)

wg_long <- wg_admins %>%
  group_by(original_id, mom_ed) %>% 
  count %>%
  filter(n == 2,  !is.na(mom_ed))

wg_long_admins <- left_join(wg_long, wg_admins)

wg_items <- get_item_data(language = "English (American)", 
                          form = "WS")
  
wg_all <- get_instrument_data(language = "English (American)", 
                                     form = "WS", 
                              items = filter(wg_items, 
                                             type == "word") %>%
                                pull(item_id))

wg_long_byitem <- filter(wg_all, data_id %in% wg_long_admins$data_id) 
```

Make main dataframe.

```{r}
d <- left_join(wg_long_byitem, 
               select(wg_long_admins, 
                      data_id, original_id, age, mom_ed, sex)) %>%
  mutate(produces = as.numeric(value == "produces")) %>%
  select(-value) %>%
  left_join(select(wg_items, num_item_id, definition, category)) %>%
  rename(person = original_id, 
         item = definition) %>%
  select(person, age, sex, mom_ed, category, item, produces)
```

## Subsampling 

```{r}
ms <- d %>%
  group_by(person, age, sex) %>%
  summarise(produces = sum(produces))

ggplot(ms, 
       aes(x = age, y = produces, group = person, col = sex)) + 
  geom_jitter(alpha = .2, width = .2)+ 
  geom_line(alpha = .2) +
  geom_smooth(aes(group = sex), span = 2, se = FALSE) + 
  ggthemes::scale_color_solarized()
  
```


Break into two time periods and subsample.

```{r}
earlier <- d %>%
  group_by(person, age) %>%
  count() %>%
  group_by(person) %>%
  mutate(earlier = ifelse(age == age[1], TRUE, FALSE),
         twos = ifelse(age[1] == 24, TRUE, FALSE)) %>%
  select(-n)

d <- left_join(d, earlier)

d_sub <- filter(d, 
                category == "food_drink", 
                twos)

d1_sub <- filter(d_sub, earlier)
d2_sub <- filter(d_sub, !earlier)
```
Plotting the subsampled data. 

```{r}
ms <- d_sub %>%
  group_by(person, age, sex) %>%
  summarise(produces = sum(produces))

ggplot(ms, 
       aes(x = age, y = produces, group = person, col = sex)) + 
  geom_jitter(alpha = .2, width = .2)+ 
  geom_line(alpha = .2) +
  geom_smooth(aes(group = sex), span = 2, se = FALSE) + 
  ggthemes::scale_color_solarized() +
  langcog::theme_mikabr() 
  
```



# Models

## 1PL with no sex differences

First fit 1pl to time 1. 

```{r}
formula_1pl <- produces ~ 1 + (1|person) + (1|item)
  
prior_1pl <-
  prior("normal(0, 3)", class = "sd", group = "person") + 
  prior("normal(0, 3)", class = "sd", group = "item")

mod_d1_1pl <- brm(formula = formula_1pl,
              prior = prior_1pl, 
              family = brmsfamily(family = "bernoulli", link = "logit"),
              data = d1_sub)
```

Basic summary of the model.

```{r}
summary(mod_d1_1pl)
```

Examine parameters. 

```{r}
plot(mod_d1_1pl)
```

Plot items. 

```{r}
items_1pl <- as_tibble(coef(mod_d1_1pl)$item[, , "Intercept"]) %>%
  mutate(item = rownames(coef(mod_d1_1pl)$item[, , "Intercept"]), 
         item = fct_reorder(item, Estimate))

ggplot(items_1pl, 
       aes(x = item, y = Estimate)) + 
  geom_pointrange(aes(ymin = Q2.5, ymax = Q97.5)) + 
  coord_flip()
```

Now plot people. 

```{r}
persons_1pl <- as_tibble(ranef(mod_d1_1pl)$person[, , "Intercept"]) %>%
  mutate(person = rownames(ranef(mod_d1_1pl)$person[, , "Intercept"])) %>%
  left_join(d1_sub %>%
              group_by(person, age, sex) %>% count()) %>%
  mutate(person = fct_reorder(person, Estimate))

ggplot(persons_1pl, 
       aes(x = person, y = Estimate, col = sex)) + 
  geom_pointrange(aes(ymin = Q2.5, ymax = Q97.5)) + 
  coord_flip()
```

## 1PL with sex differences and DIF

Now think about modeling sex differences. 

Classically, DIF would be written as a fied effect of sex plus a random effect of sex by item. 

```{r}
formula_1pl_dif <- produces ~ sex + (1 | person) + (sex | item)
  
prior_1pl_dif <-
  prior("normal(0, 3)", class = "sd", group = "person") + 
  prior("normal(0, 3)", class = "sd", group = "item") 

mod_d1_1pl_dif <- brm(formula = formula_1pl_dif,
                      prior = prior_1pl_dif, 
                      family = brmsfamily(family = "bernoulli", 
                                          link = "logit"),
                      data = d1_sub)
```
```{r}
plot(mod_d1_1pl_dif, pars = c("b_sexMale", "sd_item__sexMale"))
```


Plot items and DIF scores. Spoiler, there are no DIF effects on our subsample, food items.  

```{r}
items_1pl_dif <- bind_rows(
  as_tibble(coef(mod_d1_1pl_dif)$item[, , "Intercept"]) %>% 
    mutate(coef = "Intercept"), 
  as_tibble(ranef(mod_d1_1pl_dif)$item[, , "sexMale"]) %>% 
    mutate(coef = "sexMale")) %>%
  mutate(item = rep(rownames(coef(mod_d1_1pl_dif)$item[, , "Intercept"]),2), 
         item = fct_reorder(item, Estimate))

ggplot(items_1pl_dif, 
       aes(x = item, y = Estimate)) + 
  geom_pointrange(aes(ymin = Q2.5, ymax = Q97.5)) + 
  coord_flip() + 
  facet_grid(.~coef, scales = "free_x")
```

Now plot people. 

```{r}

persons_1pl_dif <- as_tibble(ranef(mod_d1_1pl_dif)$person[, , "Intercept"]) %>%
  mutate(person = rownames(ranef(mod_d1_1pl_dif)$person[, , "Intercept"])) %>%
  left_join(d1_sub %>%
              group_by(person, age, sex) %>% count()) %>%
  mutate(person = fct_reorder(person, Estimate))

persons_1pl_dif_coef <- as_tibble(coef(mod_d1_1pl_dif)$person[, , "Intercept"]) %>%
  mutate(person = rownames(coef(mod_d1_1pl_dif)$person[, , "Intercept"])) %>%
  left_join(d1_sub %>%
              group_by(person, age, sex) %>% count()) %>%
  mutate(person = fct_reorder(person, Estimate))

ggplot(persons_1pl_dif, 
       aes(x = person, y = Estimate, col = sex)) + 
  geom_pointrange(aes(ymin = Q2.5, ymax = Q97.5)) + 
  coord_flip()
```

## Comparison of DIF and non-DIF models

Plot sex-DIF vs. non-sex-DIF estimates for people, showing the differences in the random ability estimates.  


```{r}
compare_dif <- tibble(estimate = persons_1pl$Estimate, 
                      estimate_dif = persons_1pl_dif$Estimate,
                      estimate_dif_coef = persons_1pl_dif_coef$Estimate,
                      sex = persons_1pl$sex)

ggplot(data = compare_dif, 
       aes(x = estimate, y = estimate_dif, 
           col = sex)) +
  geom_point() + 
  geom_abline(lty = 2) 

# p2 <- ggplot(data = compare_dif, 
#        aes(x = estimate, y = estimate_dif_coef, 
#            col = sex)) +
#   geom_point() + 
#   geom_abline(lty = 2) + 
#   theme(legend.pos = "bottom")
# 
# cowplot::plot_grid(p1,p2)
```


You can get:

1. Estimate of absolute ability. How good are you compared to other people. 
2. Demographically-adjusted estimate of ability. Effectively, how good are are you compared to other people, if you weren't in your demographic group. 

This second is of interest in reasoning about "ability in context." 

One way of formalizing the affirmative action hypothesis is the idea that, for some demographics, (2) should be a better predictor of future performance than (1). 

<!-- # More parameters  -->

<!-- (too slow at the moment) -->

<!-- ```{r, eval=FALSE} -->
<!-- formula_2pl <- bf(produces ~ exp(logalpha) * eta, -->
<!--   eta ~ 1 + (1 |i| item) + (1 | person),  -->
<!--   logalpha ~ 1 + (1 |i| item), -->
<!--   nl = TRUE) -->

<!-- prior_2pl <- -->
<!--   prior("normal(0, 5)", class = "b", nlpar = "eta") + -->
<!--   prior("normal(0, 1)", class = "b", nlpar = "logalpha") + -->
<!--   prior("normal(0, 3)", class = "sd", group = "person", nlpar = "eta") + -->
<!--   prior("normal(0, 3)", class = "sd", group = "item", nlpar = "eta") + -->
<!--   prior("normal(0, 1)", class = "sd", group = "item", nlpar = "logalpha") -->


<!-- mod_d1 <- brm(formula = formula_2pl, -->
<!--               prior = prior_2pl,  -->
<!--               family = brmsfamily(family = "bernoulli", link = "logit"), -->
<!--               data = d1) -->

<!-- ``` -->

